A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-06-09 trunk/textos/optpaxos/draftv3.tex
shit's getting complete

We define $sendersTo(G)$ as the set of groups which are able to send a message to $G$. Also, we define $receiversFrom(G)$ as the set of groups who are able to receive a message from $G$.

@@

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption A1, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the messages with timestamp up to $t$ can be delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

@@

The problem with Algorithm \ref{algorithm:deliveryminimal} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the delivery of new messages. However, liveness can be easily provided by sending periodic empty messages from $G$ to each $G' \in receiversFrom(G)$ to which no message has been sent for a specified time period. Section \ref{sec:nullperiodic} describes such approach. In section \ref{sec:nullondemand}, another approach, based on sending empty messages when requested, is presented. The trade-off between them concerns message delivery latency against number of control messages sent.

@@

There is a way to provide liveness with a lower delay for delivering messages, although that would imply creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order for some group to deliver $m$. More formally, \mbox{$blockers(m) = \{G_B \neq m.src : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$}. The idea is that, once each group $G_B$ in $blockers(m)$ has sent\footnote{Here, by `sending', we mean that the message has been received at the destination.} a barrier $b > m.ts$ to all the groups belonging to \mbox{$m.dst \cap receiversFrom(G_B)$}, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ has a message $m$ to send, $p$ sends $m$ to every $p' \in m.dst \cup G \cup blockers(m)$. It is sent to the groups in $blockers(m)$, so that they know that there is a message whose delivery will be blocked until they send a proper barrier to unblock it. Finally, a group never blocks the delivery of its own messages -- hence the condition $G_B \neq m.src$ -- because the message itself can be seen as a barrier from its source group.

When the process $p'$ of some group $G'$ receives $m$, such that $G' \in blockers(m)$, $p'$ knows that there might be other groups depeding on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its $optPending$ queue -- to ensure that, once it goes to $propPending$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the $null$ message will be proposed in $G'$ and, once $null$ is decided, each process of $G'$ will send a $\{null, \text{`cons'}\}$ message to each process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

>>>>

We define $sendersTo(G)$ as the set of other groups which are able to send a message to $G$. Also, we define $receiversFrom(G)$ as the set of other groups who are able to receive a message from $G$.

@@

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption A1, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' \neq G : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the messages with timestamp up to $t$ can be delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

@@

The problem with Algorithm \ref{algorithm:deliveryminimal} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the delivery of new messages. However, liveness can be easily provided by sending periodic empty messages from $G$ to each $G' \in receiversFrom(G)$ to which no message has been sent for a specified time period. Section \ref{sec:nullperiodic} describes such approach. In section \ref{sec:nullondemand}, another approach, based on sending empty messages when requested, is presented. The trade-off between them concerns message delivery latency against number of control messages sent. In either case, the empty messages are handled as ordinary messages, except they are never delivered to the application.

@@

There is a way to provide liveness with a lower delay for delivering messages, although that would imply creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order for some group to deliver $m$. More formally, \mbox{$blockers(m) = \{G_B \neq m.src : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$}. The idea is that, once each group $G_B$ in $blockers(m)$ has sent\footnote{Here, by `sending', we mean that the message has been received at the destination.} a barrier $b > m.ts$ to all the groups belonging to \mbox{$m.dst \cap receiversFrom(G_B)$}, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ knows that a message $m$ has just been decided in some consensus instance within $G$, $p$ requests a barrier to every $p' \in blockers(m)$. Such request is sent to the processes in the groups inside $blockers(m)$, so that they know that there is a message whose delivery will be blocked until they send a proper barrier to unblock it. Finally, a group never blocks the delivery of its own messages -- hence the condition $G_B \neq m.src$ -- because the message itself can be seen as a barrier from its source group.

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$propPending$ queue. % -- to ensure that, once it goes to $propPending$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

\blankline
\INDENT{\textit{When} \rmdel{$m'$}}
  \IF{$G = m'.src$}
    \STATE $propPending \leftarrow propPending \cup \{m'\}$
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $barPending \leftarrow barPending \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $propPending \leftarrow propPending \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When} Decide($k$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$
    \ENDIF
    \STATE $decided \leftarrow decided \cup \{m\}$
    \IF{$m \neq null$}
      \STATE \rmcast{$m$}{$(m.dst \setminus \{G\}) \cup blockers(m)$} \COMMENT{the message is sent to blockers(m), as a barrier request}
    \ELSE
      \STATE \rmcast{$m$}{$m.dst$} \COMMENT{no need to unblock $null$, besides it would create infinite messages}
    \ENDIF
  \ENDINDENT
  \STATE $nextProp \leftarrow k + 1$
  \STATE $k \leftarrow k + 1$
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$ }}
  \STATE $barPending \leftarrow barPending \setminus \{m\}$
  \IF{$m \neq null$}
    \STATE \consdel{$m$}
  \ENDIF
  \STATE $delivered \leftarrow delivered \cup \{m\}$

\caption{\cmsend{m} requesting empty messages -- executed by every process $p$ from group $G$}
\label{algorithm:nullondemand}
\end{distribalgo}
\end{algorithm}
